---
title: "Assignment"
author: "Baris Uzel"
date: "2023-11-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment
```{r}

#load packages
library(raster)
library(rasterVis)
library(ncdf4)
library(dplyr)
library(lubridate)
library(ggplot2)
library(animation)
```

## Data Import & Preparation

```{r}
#Import environmental data - Please download all the data into a folder called data and set working the directory outside the data folder.  

tempdata <- brick("data/2mtemp.nc") #Import temperature data
evodata <- brick("data/totalevaporation.nc") #Import total evaporation data
precipdata <-brick("data/totalprecipitation.nc") #Import total precipitation data
socdata <- raster("data/galicia_soc.tif") # Import soil organic carbon stock data
landdata <- raster("data/lccglobal.nc") # Import global land cover classification data


frame <- extent(c(-8.3, -7.3, 41.8, 42.8)) # Coordinates for cropping extent

gal_soc <- crop(socdata, extent(frame))
gal_land <- crop(landdata, extent(frame))
gal_precip <- crop(precipdata, extent(frame))
gal_temp <- crop(tempdata, extent(frame))
gal_evo <- crop(evodata, extent(frame))

```
## Reprojection

```{r}

# Assign minimum resolution
resolutions <- c(res(gal_precip), res(gal_temp), res(gal_evo), res(gal_soc), res(gal_land))
min_resolution <- min(resolutions) # determine minimum resolution
projection <- crs(gal_temp) # Define collective coordinate system
crs(gal_land) <-projection #landclassification data missing appropriate projection

# Reproject using coordinate system and minimum resolution
projection_temp <- projectRaster(gal_temp, crs = projection, method = "ngb", res = min_resolution, block.size =c(25,25))
projection_precip <- projectRaster(gal_precip, crs = projection, method = "ngb", res = min_resolution, block.size =c(25,25))
projection_evo <- projectRaster(gal_evo, crs = projection, method = "ngb", res = min_resolution, block.size =c(25,25))
projection_soc <- projectRaster(gal_soc, crs = projection, method = "ngb", res = min_resolution, block.size =c(25,25))
projection_land <- projectRaster(gal_land, crs = projection, method = "ngb", res = min_resolution, block.size =c(25,25))

```

#Creating basic maps to explore data
```{r}
scsmap <- levelplot(projection_soc, margin = FALSE, #mapping soil carbon stocks
          main = "Soil Carbon Stocks of Galicia, Spain (t/ha)",
          col.regions = colorRampPalette(c("light green", "dark green"))(20))

png("galicia_scs.png", width = 800, height = 600)   #Save soil carbon stock map to our working                                                           directory
scsmap
dev.off()

# RGB values for each land cover class according to CDS Documentation
lcc_rgb <- matrix(c(
  0, 0, 0,   # No Data
  255, 255, 0,   # Cropland, rainfed
  170, 240, 240,   # Cropland, irrigated or post-flooding
  220, 240, 100,   # Mosaic cropland / natural vegetation
  200, 200, 100,   # Mosaic natural vegetation / cropland
  0, 100, 0,   # Tree cover, broadleaved, evergreen
  0, 160, 0,   # Tree cover, broadleaved, deciduous
  0, 60, 0,   # Tree cover, needleleaved, evergreen
  40, 80, 0,   # Tree cover, needleleaved, deciduous
  120, 130, 0,   # Tree cover, mixed leaf type
  140, 160, 0,   # Mosaic tree and shrub / herbaceous cover
  190, 150, 0,   # Mosaic herbaceous cover / tree and shrub
  150,  100, 0,   # Shrubland
  255, 180, 50,   # Grassland
  255, 220, 210,   # Lichens and mosses
  255, 235, 175,   # Sparse vegetation
  0, 120, 90,   # Tree cover, flooded, fresh or brackish water
  0, 150, 120,   # Tree cover, flooded, saline water
  0, 220, 130,   # Shrub or herbaceous cover, flooded
  195, 20,   0,   # Urban areas
  255, 245, 215,   # Bare areas
    0,   70, 200,   # Water bodies
  255, 255, 255    # Permanent snow and ice
), ncol = 3, byrow = TRUE)


lcc_colours <- apply(lcc_rgb, 1, function(rgb) rgb(rgb[1], rgb[2], rgb[3], maxColorValue = 255)) #Assign RGB values to land cover classes
lcc_colourmap <- setNames(lcc_colours, seq(0, 220, by = 10)) # Create vector to map land cover codes to RGB colors


lccplot <-  levelplot(   # Plot Land cover classification
    projection_land,  
    margin = FALSE, 
    main = "Land Cover Classification of Galicia, Spain", 
    col.regions = lcc_colourmap,
    at = seq(0, 220, length.out = nrow(lcc_rgb) + 1), 
    colorkey = list(
      labels = list(
        at = seq(0, 220, length.out = nrow(lcc_rgb)),
        labels = c(
        "No Data", "Cropland, rainfed", "Cropland, irrigated or post-flooding", 
        "Mosaic cropland / natural vegetation", "Mosaic natural vegetation / cropland",
        "Tree cover, broadleaved, evergreen", "Tree cover, broadleaved, deciduous", 
        "Tree cover, needleleaved, evergreen", "Tree cover, needleleaved, deciduous",
        "Tree cover, mixed leaf type", "Mosaic tree and shrub / herbaceous cover",
        "Mosaic herbaceous cover / tree and shrub", "Shrubland", "Grassland", 
        "Lichens and mosses", "Sparse vegetation",
        "Tree cover, flooded, fresh or brackish water", 
        "Tree cover, flooded, saline water", 
        "Shrub or herbaceous cover, flooded", "Urban areas", 
        "Bare areas", "Water bodies", "Permanent snow and ice")),
    width = 1,  
    height = 1,  
    space = "right"))

png("galicia_lcc.png", width = 800, height = 600)   #Save land cover classification map
lccplot
dev.off()
```
#Create a Time-series animation for visualization

```{r}
years <- 2000:2022 #define year range


# Below are loops that create calendar plots of every year between 2000:2022 and save them as the environmental variable+calendar year. The three loops do the same function for Evapotranspiration, Temperature, and Precipitation respectively. 

for (year in years) {                       #Evapotranspiration
  start_layer <- (year - 2000) * 12 + 1
  end_layer <- start_layer + 11
  
  current_months <- rep(month.abb, length.out = end_layer - start_layer + 1)
  
  current_evaporation <- levelplot(
    projection_evo,
    layers = start_layer:end_layer,
    main = paste("Monthly Average Evapotranspiration for the year", year,
                 "(m of water eq.)"),
    names.attr = current_months,
    colorkey = list(
      width = 1,
      height = 1,
      space = "right"
    )
  )
  
  assign(paste0("Evaporation", year), current_evaporation)
}

for (year in years) {                   #Temperature
  start_layer <- (year - 2000) * 12 + 1
  end_layer <- start_layer + 11
  
  current_months <- rep(month.abb, length.out = end_layer - start_layer + 1)
  
  current_temperature <- levelplot(
    projection_temp,
    layers = start_layer:end_layer,
    main = paste("Monthly Average Temperature for the year", year,"(K)"),
    names.attr = current_months,
    colorkey = list(
      width = 1,
      height = 1,
      space = "right"
    )
  )
  
  assign(paste0("Temperature", year), current_temperature)
}

for (year in years) {                         #Precipitation
  start_layer <- (year - 2000) * 12 + 1
  end_layer <- start_layer + 11
  
  current_months <- rep(month.abb, length.out = end_layer - start_layer + 1)
  
  current_precipitation <- levelplot(
    projection_precip,
    layers = start_layer:end_layer,
    main = paste("Monthly Average Precipitation for the year", year,"(m)"),
    names.attr = current_months,
    colorkey = list(
      width = 1,
      height = 1,
      space = "right"
    )
  )
  
  assign(paste0("Precipitation", year), current_precipitation)
}

## Create Time Series Animations.
#The 3 loops below create time series animations of rasters. This is a good supplementary way to visualize the change in evaporation/temperature/precipitation over the map.  
saveGIF({
  for (year in years) {
    current_plot <- get(paste0("Evaporation",year))
    print(current_plot)
  }
}, movie.name = "Evaporation_timseries.gif", ani.width = 800, ani.height = 600, interval = 1)


saveGIF({
  for (year in years) {
    current_plot <- get(paste0("Temperature",year))
    print(current_plot)
  }
}, movie.name = "Temperature_timseries.gif", ani.width = 800, ani.height = 600, interval = 1)


saveGIF({
  for (year in years) {
    current_plot <- get(paste0("Precipitation",year))
    print(current_plot)
  }
}, movie.name = "Precipitation_timseries.gif", ani.width = 800, ani.height = 600, interval = 1)

```
# My initial strategy for joint analysis was to:
#     1. Overay land cover classification raster on to the temp, precip, evo rasterBricks
#     2. Extract the temp/precip/evo/ values for each land cover class.
#     3. Take the mean for every variable for every land cover at every month of year.
#     4. Create a 'traditional' timeseries to visualize trends.
#     5. Chose landcover/variable combinations of interest

#This approach to 'traditional timeseries' and spatial statistics proved to be computationally demanding. The results of my first loop (see code chunk below) took a very long time to process and after an overnight run attempt, I chose to continue my analysis with a different format..

```{r}

temp_results <- list()   # An empty list to store the results
custom_function <- function(x, y) {     #mean
  mean(x, na.rm = TRUE)
}

for (i in 1:nlayers(projection_temp)) {
  result_layer <- raster::overlay(projection_temp[[i]], projection_land, fun = custom_function)
  temp_results[[i]] <- result_layer
}
templandoverlay <- brick(temp_results)
print(temp_results)

```


# Instead of loading my data as rasters, I am going to load it as NetCDF format and continue 'traditional time-series' and spatial statistics in this format. 

```{r}
#Load Data
tempnc <- nc_open("data/2mtemp.nc")
evonc <- nc_open("data/totalevaporation.nc")
precipnc <-nc_open("data/totalprecipitation.nc")
globallandnc <- nc_open("data/lccglobal.nc")

#Extract lat,long,time
lat <- ncvar_get(tempnc, "latitude")
lon <- ncvar_get(tempnc, "longitude")
time <-ncvar_get(tempnc, "time") 
time <- as.POSIXct(time * 3600, origin = "1900-01-01", tz = "GMT")
lonlattime <- as.matrix(expand.grid(lon,lat,time))

#Extract variables and store in vector form
temp_array <- ncvar_get(tempnc, "t2m")
evo_array <- ncvar_get(evonc, "e")
precip_array <- ncvar_get(precipnc, "tp")
temp_vec <- as.vector(temp_array)
evo_vec <- as.vector(evo_array)
precip_vec <- as.vector(precip_array)


#Build dataframe
envdata <- data.frame(cbind(lonlattime, temp_vec, evo_vec, precip_vec))
colnames(envdata) <- c("Lon","Lat","Date","Temp", "Evo", "Precip")
envdata$Date <- as.Date(envdata$Date)
envdata$Temp <- as.numeric(envdata$Temp)
envdata$Evo <- as.numeric(envdata$Evo)
envdata$Precip <- as.numeric(envdata$Precip)
print(envdata)

#Make Temperature Time Series
tempseries <- envdata %>%
                group_by(Date) %>%
                summarize(Temp = mean(Temp, na.rm = TRUE))  %>%
                  mutate(Temp.C = Temp-273.15)

tempseries <- tempseries %>%
  ggplot(aes(x=Date, y=Temp.C)) +
  geom_point(size=0.2, color="black") +
  geom_line(size=0.5, color="black") +
  labs(title = "Monthly average temperature (°C) in Galicia, Spain between 2000-2022",
       x = "Year",
       y = "Temperature (°C)") +
  theme_classic()
tempseries

#Evaporation Time Series
evoseries <- envdata %>%
                group_by(Date) %>%
                summarize(Evo = mean(Evo, na.rm = TRUE)) %>%
                  mutate(Evo = abs(Evo))


evoseries <- evoseries %>%
  ggplot(aes(x=Date, y=Evo)) +
  geom_point(size=0.2, color="black") +
  geom_line(size=0.5, color="black") +
  labs(title = "Monthly average evaporation in Galicia, Spain between 2000-2022",
       x = "Year",
       y = "Evapotranspiration (m water eq.)") +
  theme_classic()
evoseries


#Precipitation Time Series
precipseries <- envdata %>%
                group_by(Date) %>%
                summarize(Precip = mean(Precip, na.rm = TRUE))

precipseries <- precipseries %>%
  ggplot(aes(x=Date, y=Precip)) +
  geom_point(size=0.2, color="black") +
  geom_line(size=0.5, color="black") +
  labs(title = "Monthly average precipitation in Galicia, Spain between 2000-2022",
       x = "Year",
       y = "Precipitation (m)") +
  theme_classic()

precipseries
```

#Now I want to investigate if things have changed BY landcover
#To do that I need to crop and interpolate the globallandcover classificationt data that has a much finer spatial resolution. 
```{r}
#Load Data
globallandnc <- nc_open("data/lccglobal.nc") #different spatial resolution
tempnc <- nc_open("data/2mtemp.nc")


#Extract lat & long
landlat <- ncvar_get(globallandnc, "lat")
landlon <- ncvar_get(globallandnc, "lon")
landlat<- subset(landlat, landlat >= 41.8 & landlat <= 42.8) #crop to frame
landlon <- subset(landlon, landlon >= -8.3 & landlon <= -7.3)
lonlat <- as.matrix(expand.grid(landlon,landlat))

#Extract variables based on lat lon
lat <- ncvar_get(tempnc, "latitude")
lon <- ncvar_get(tempnc, "longitude")

lat_min <- min(lat)
lat_max <- max(lat)
lon_min <- min(lon)
lon_max <- max(lon)

#Extract landcover classifications within the coordinate range
land_array <- ncvar_get(globallandnc, "lccs_class", 
                       start = c(which(globallandnc$dim$lon$vals >= lon_min & globallandnc$dim$lon$vals <= lon_max)[1], 
                                 which(globallandnc$dim$lat$vals >= lat_min & globallandnc$dim$lat$vals <= lat_max)[1], 
                                 1), 
                       count = c(length(which(globallandnc$dim$lon$vals >= lon_min & globallandnc$dim$lon$vals <= lon_max)), 
                                 length(which(globallandnc$dim$lat$vals >= lat_min & globallandnc$dim$lat$vals <= lat_max)), 
                                 1, ncol(globallandnc)))

#Build the dataframe
land_vec <- as.vector(land_array)
landcoverdata <- data.frame(cbind(lonlat, land_vec))
colnames(landcoverdata) <- c("Lon", "Lat", "landcover")


##Interpolation - Below I interpolate the fine land cover classification data to the same spatial resolution as the coarser environmental variable data. 

#I define the spatial resolution
lon_values <- seq(-7.3, -8.3, by = -0.1)
lat_values <- seq(42.8, 41.8, by = -0.1)

s_resolution <- expand.grid(lon = lon_values, lat = lat_values)

#Define the interpolation function. The method I chose here is to count all landcover classifications in +and- 0.1 lat/long range of the point and assign the value of the most common (or most repeating) classification within that range. 
function_interpolation <- function(lon_start, lon_end, lat_start, lat_end, data) {
  data %>%
    filter(between(Lon, lon_start, lon_end) & between(Lat, lat_start, lat_end)) %>%
    group_by(landcover) %>%
    summarize(count = n()) %>%
    arrange(desc(count)) %>%
    slice(1) %>%
    pull(landcover)
}
#perform interpolation
interpolation <- s_resolution %>%
  rowwise() %>%
  mutate(landcover = function_interpolation(lon - 0.1, lon + 0.1, lat - 0.1, lat + 0.1, landcoverdata))

#check results
print(interpolation)

#Merge datasets
colnames(interpolation) <- c("Lon", "Lat", "landcover")
all_data <- merge(envdata, interpolation, by = c("Lon", "Lat"))

##Now that we have all datasets in one folder, we can make the categorical timeseries

#add classification to data
all_data <- all_data %>%
  mutate(classification = case_when(
    landcover == 10 ~ "Cropland, rainfed",
    landcover == 40 ~ "Mosaic natural vegetation / cropland",
    landcover == 60 ~ "Tree cover, broadleaved, deciduous",
    landcover == 70 ~ "Tree cover, needleleaved, evergreen",
    landcover == 100 ~ "Mosaic tree and shrub / herbaceous cover",
    landcover == 130 ~ "Grassland",
  ))

#group temperature, evapotranspiration, and precipitation by classification
all_data_temp <- all_data %>%
  group_by(classification, Date) %>%
  summarize(mean_temp = mean(Temp))

all_data_evo <- all_data %>%
  group_by(classification, Date) %>%
  summarize(mean_evo = mean(Evo))

all_data_precip <- all_data %>%
  group_by(classification, Date) %>%
  summarize(mean_precip = mean(Precip))

#Unit Conversion
all_data_temp <- all_data_temp %>%
                  mutate(Temp.C = mean_temp-273.15)
all_data_temp
#Define Standard Colours as per CDS documentation
class_colors <- c(
  "Cropland, rainfed" = rgb(255, 255, 0, maxColorValue = 255),
  "Mosaic natural vegetation / cropland" = rgb(200, 200, 100, maxColorValue = 255),
  "Tree cover, broadleaved, deciduous" = rgb(0, 160, 0, maxColorValue = 255),
  "Tree cover, needleleaved, evergreen" = rgb(0, 60, 0, maxColorValue = 255),
  "Mosaic tree and shrub / herbaceous cover" = rgb(140, 160, 0, maxColorValue = 255),
  "Grassland" = rgb(255, 180, 50, maxColorValue = 255)
)

#Plot
tempbyclass <- ggplot(all_data_temp, aes(x = Date, y = Temp.C, color = classification, group = classification)) +
  geom_line() +
  labs(title = "Temperature Trends by Landcover Classification",
       x = "Date",
       y = "Mean Temperature (°C)") +
  theme_classic() +
    theme(legend.key.size = unit(0.1, "cm"),
        legend.position = "bottom") +
  guides(color = guide_legend(keywidth = 0.8, keyheight = 0.5, title.position = "top"))

png("temperature_timeseries_byclass.png", width = 800, height = 600)  #Save
tempbyclass
dev.off()

evobyclass <- ggplot(all_data_evo, aes(x = Date, y = mean_evo, color = classification, group = classification)) +
  geom_line() +
  labs(title = "Evaporation Trends by Landcover Classification",
       x = "Date",
       y = "Mean Evapotranspiration (m water eq.)") +
  theme_classic() +
    theme(legend.key.size = unit(0.1, "cm"),
        legend.position = "bottom") +
  guides(color = guide_legend(keywidth = 0.8, keyheight = 0.5, title.position = "top"))

png("evapotranspiration_timeseries_byclass", width = 800, height = 600)  #Save
evobyclass
dev.off()

precipbyclass <- ggplot(all_data_precip, aes(x = Date, y = mean_precip, color = classification, group = classification)) +
  geom_line() +
  labs(title = "Precipitation Trends by Landcover Classification",
       x = "Date",
       y = "Mean Precipitation (m)") +
  theme_classic() +
    theme(legend.key.size = unit(0.1, "cm"),
        legend.position = "bottom") +
  guides(color = guide_legend(keywidth = 0.8, keyheight = 0.5, title.position = "top"))

png("precipitation_timeseries_byclass.png", width = 800, height = 600)  #Save
precipbyclass
dev.off()

```

#Sampling Design
```{r}

soc_values <- extract(gal_soc, frame) #extract SOC values
print(mean(soc_values)) # determine mean
print(sd(soc_values))   # determine the standard deviation

# 95% confidence level = Z score 1.96
Zsq <- 1.96^2
#Standard deviation squared
Sdsq <- 14.8^2
# 10% of mean for margin of error
E <- (0.1*81)^2

#Finalize equation
sample_size <- (Zsq*Sdsq/E)
print(sample_size) # 13 samples would have to be taken in the region to have 95% CI


## Final Comments on Sampling

#If I was to create a sampling plan in the region (or any region for that matter), my goal would be to account for as much soil variability as possible with enough repetitions. To achieve this, I would take additional soil data such as moisture and elevation to be able to capture the variability in low/high moisture soils, low/high elevation soils, and so on. I would build a model to generate specific sampling points on locations that represent this variability. Of course, in that case I would revise the above calculation and the samples necessary would increase to have 95%CI.
```